\section*{Problem 2: Locality-Sensitive Hashing (LSH) for Angle Similarity}

Suppose our set of $n$ points $D = \left\{p_1,\ldots,p_n\right\}$ are vectors in
$d$ dimensions. Our problem is: given a query point $q$ find a point $p \in D$,
which has a small angle with $q$. Recall that the angle between two vectors $a$
and $b$ is
$\cos^{-1}\left(\frac{a \cdot b}{\lVert a \rVert \lVert b \rVert} \right).$

As doing this exactly may be computationally expensive, let us try to do this
approximately with a fast algorithm. The approximate objective is as follows:
suppose there exists a point $p \in D$ which has angle less than $\theta$ with
$p$, then our goal is return a point with angle less than $c\theta$, where
$c > 1$.

Let us try to do this with LSH. Let us consider the a family of hash functions,
where $h(p) = \operatorname{sign}\left(u \cdot p\right)$ where we will sample
$u$ uniformly at random from a Gaussian (or from a unit sphere).

\begin{enumerate}
\item Provide an exact expression for
  $\mathbb{P}\left(h\left(p\right) = h\left(p^\prime\right)\right)$ based on
  some geometric relation between $p$ and $p^\prime$.

  \subsection*{Solution}

  Define
  \begin{equation}
    \operatorname{angle}\left(u, v\right) =
    \cos^{-1}\left(\frac{u \cdot v}{\lVert u \rVert \lVert v \rVert} \right),
  \end{equation}
  which is the angle between two vectors.

  Then,
  \begin{equation}
    \boxed{\mathbb{P}\left(h\left(p\right) = h\left(p^\prime\right)\right)
      = 1 - \frac{\operatorname{angle}\left(p,p^\prime\right)}{\pi}.}
  \end{equation}
\item Provide an expression for $P_1$ and $P_2$ in terms of $\theta$ and
  $c\theta$. Note that since we want a small angle, we should use:
  \begin{enumerate}
  \item If $\operatorname{angle}\left(p,p^\prime\right) < \theta$, then
    $\mathbb{P}\left(h\left(p\right) = h\left(p^\prime\right)\right) \geq P_1$.        
  \item If $\operatorname{angle}\left(p,p^\prime\right) > c\theta$, then
    $\mathbb{P}\left(h\left(p\right) = h\left(p^\prime\right)\right) \leq P_2$.
  \end{enumerate}

  \subsection*{Solution}

  If $\operatorname{angle}\left(p,p^\prime\right) < \theta$, then
  \begin{align*}
    \mathbb{P}\left(h\left(p\right) = h\left(p^\prime\right)\right)
    &= 1 - \frac{\operatorname{angle}\left(p,p^\prime\right)}{\pi} 
    \geq 1 - \frac{\theta}{\pi},
  \end{align*}
  so $\displaystyle\boxed{P_1 = 1 - \frac{\theta}{\pi}.}$

  If $\operatorname{angle}\left(p,p^\prime\right) > c\theta$, then
  \begin{align*}
    \mathbb{P}\left(h\left(p\right) = h\left(p^\prime\right)\right)
    &= 1 - \frac{\operatorname{angle}\left(p,p^\prime\right)}{\pi}
    \leq 1 - \frac{c\theta}{\pi},
  \end{align*}
  so $\displaystyle\boxed{P_2 = 1 - \frac{c\theta}{\pi}.}$
\item Provide expressions for query time for point $q$, the space to store the
  hash tables, and the construction time of our datastructure.

  \subsection*{Solution}

  Suppose we have $L$ hash functions. If we use the algorithm discussed in
  class, to query a point, we need to compute $L$ hashes. Then, up to $3$ times,
  we iterate through the buckets: for each bucket, we choose a point and check
  how close it is to $q$; if it is $c\theta$ close, we stop. The worst case is
  that we decide there exists no point that is $\theta$ close to $q$. In this
  case, we iterate through the $L$ buckets $3$ times, so the time complexity is
  $O(L)$.
  
  For the space needed to store the hash tables, we need to store $L$ bits for
  each point, so the space needed is $O(nL)$
  
  For construction, we need to compute $L$ hashes for each point, so the
  computational complexity is $O(nL)$ as well.  
\end{enumerate}
